---
title: "Super Bowl Ads"
output: 
  github_document:
    toc: true
    toc_depth: 3
---
## Background
This notebook focuses on super bowl ads data! Again, this data is retrieved
from the `tidy-tuesday` collection. 

To give a general overview, this dataset was curated by the team at 
[FiveThirtyEight](https://fivethirtyeight.com/). Per their [article](https://projects.fivethirtyeight.com/super-bowl-ads/) (which I will
refrain from reading to have a true explanatory experience), they describe how
they composed this dataset:

> We watched 233 ads from the 10 brands that aired the most spots in all 
21 Super Bowls this century, according to superbowl-ads.com.1 While we watched, 
we evaluated ads using seven specific criteria, marking every spot as a "yes" 
or "no" for each.

For more information the dataset itself and
the characteristics, feel free to check it out [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-03-02/readme.md)

---

## Setup

First, let's load the libraries and data:
```{r libraries, message=FALSE}
library(tidyverse)
library(ggplot2)
library(tidytuesdayR) # library from which we'll retrieve the data
library(patchwork)
library(stringr)
library(reshape2)
library(scales)
# Let's also set a theme for our plots
theme_set(theme_minimal())
```

Now let's load the data from `tidy-tuesday` (March 2nd, 2021):
```{r message=FALSE}
super_bowl_ads <- tt_load('2021-03-02')
super_bowl_ads <- super_bowl_ads$youtube
```

---

## Exploratory Data Analysis (EDA)

To begin, I like to just take a quick gander at the data to see what it looks
like:
```{r}
head(super_bowl_ads)
```

Immediately, there are a couple of columns I'd like to drop that really won't be
useful for analysis. Those columns are:

  - `superbowl_ads_dot_com_url`
  - `youtube_url`
  - `etag`
  - `thumbnail`

I'm removing these columns since these columns are metadata that won't add any
sort of value to our analyses.
```{r}
super_bowl_ads <- subset(
  super_bowl_ads, 
  select = -c(superbowl_ads_dot_com_url,
              youtube_url,
              etag,
              thumbnail)
  )
```

Having removed those columns, let's take a quick look to see if we have any 
abnormalities like `NaN` values:
```{r}
sapply(super_bowl_ads, function(x) sum(is.na(x)))
```

We find that there are several missing values in the `kind` and `view_count`
which will make analyzing those rows impossible. So let's look at those rows 
and make sure that we can drop them.

```{r}
# Show the number of rows in the dataframe
print(nrow(super_bowl_ads))

# Let's just drop the columns which have either missing views, and like_counts
# since those are the most important from an analysis point of views
super_bowl_ads <- super_bowl_ads[!is.na(super_bowl_ads$view_count), ]
super_bowl_ads <- super_bowl_ads[!is.na(super_bowl_ads$like_count), ]

# Let's view the missing values again:
sapply(super_bowl_ads, function(x) sum(is.na(x)))
print(nrow(super_bowl_ads))
```

In total, we dropped about **22 observations**, which is not too many considering
we started with 247 observations (less than 10%).

With our data relatively clean (for analysis), let's formulate a list of 
questions we can answer regarding this dataset:

1. Are certain types of commercials associated with a specific type of brand?
How has the type of commercials changed over time?
2. What are the most popular commercials? We can determine this by the number 
of views that a video received and furthermore, we can do some feature
engineering to further define other sorts of superlatives (most controversial,
likeable, etc.)
3. Does having more categories help or hurt a commercial? Tying in with question
2, do the most popular commercials typically have two or more categories? Or are
they more focused on one single category?

Within each question, we will likely have further sub-questions, but for now, 
let's get started 🏃🏽‍♂️!

---

### (1) Ad categories over time

The categories that we're working with are:

- `funny`
- `show_product_quickly`
- `patriotic`
- `celebrity`
- `danger`
- `animals`
- `use_sex`

Some of these are more subjective than others (`funny`, `show_product_quickly`), 
but we'll trust the team at FiveThirtyEight has a good sense of humor and is 
relatively impatient 😉. 

First, let's look at who our 10 brands are 
```{r}
unique(super_bowl_ads$brand)
```

We have:

- 3 car brands
- 5 food/beverage brands
- 1 financial brand
- 1 sports brand

Before looking at the ad categories, how is the distribution of ads amongst the
brands?
```{r}
super_bowl_ads %>%
  group_by(brand) %>%
  summarize(total_ads = n()) %>%
  arrange(desc(total_ads))
```

We can already see that the number of ads are heavily skewed towards 
our food/beverage brands. So when we look at the categories of each of the
different ads, we should consider some sort of percentage to combat the 
disparity in the number of ads between each brand.

Let's go ahead and now create this grouping for our different categories:
```{r}
# Group the columns by each category
group_category_by_brand <- super_bowl_ads %>%
  group_by(brand) %>%
  summarize(
    funny = sum(funny),
    show_product_quickly = sum(show_product_quickly),
    patriotic = sum(patriotic), 
    celebrity = sum(celebrity),
    danger = sum(danger),
    animals = sum(animals),
    use_sex = sum(use_sex))

# Convert tible to dataframe
group_category_by_brand <- as.data.frame(group_category_by_brand)

group_category_by_brand
```

Having generated this table, let's make a useful visualization out of it to 
extract greater meaning from it. However, to make said visualization, we'll need
to use the `melt` function from `reshape2` to make it usable:

```{r}
# Melt all the values from the dataframe to create a new one
melt_df <- melt(group_category_by_brand, id.vars = 1)

# Generate a grouped barplot
ggplot(
  data = melt_df,
  aes(x = value, y = reorder(brand, +value))) + 
  geom_bar(
    aes(fill = variable),
    stat = "identity",
    position = "stack") + 
  scale_fill_brewer(palette="Spectral") + 
  labs(
    title = "Funny Bud Light Ads are in Abundance!",
    x = "Number of Ads",
    y = "Brand"
  )
```

This is quite a nice and informative graph. We can see that the majority of the
ads are Bud Light ones and that they always tend to be funny and show their 
product quickly. Already from this graph, we can see that the majority of the 
ads are funny ones, but let's look at the distribution of ads across the 
different groupings:

```{r}
# Get the sums of each column
sum_df <- as.data.frame(
  as.list(
    colSums(Filter(is.numeric, group_category_by_brand))
  )
)

# Transpose the dataframe and make the index a new row called category
sum_df <- t(sum_df)
sum_df <- as.data.frame(cbind(category = rownames(sum_df), sum_df))
colnames(sum_df)[2] ="count"
sum_df$count = as.numeric(sum_df$count)

# Create the bar plot
sum_df %>%
  ggplot() + 
  aes(x = reorder(category, -count), y = count) + 
  geom_bar(
    stat = "identity",
    fill = '#3388bd') + 
  labs(
    title = "Show them Quick and Make them Funny!",
    x = "Category",
    y = "Number of Ads"
  )
```

Ads which are funny and show the product quickly are the most prominent types of
ads amongst the biggest brands! From a general perspective, that is how super
bowl ads are typically thought of. 

However, while those are all nice sets of graphs, we still don't have a 
definitive answer to what types of ads do companies use. Let's answer that 
with certainty now:
```{r}
# Set up plotting space
brand_names <- as.list(group_category_by_brand$brand)

# Generate a list to house the plot
plots <- list()

for (i in brand_names) { # Loop over loop.vector
  
  # Subset the dataframe
  df_subset <- melt_df[melt_df$brand == i, ]

  # store data in column.i as x
  plot <- df_subset %>%
    ggplot() + 
    aes(x = brand, y = value) + 
    geom_bar(
      aes(fill = variable),
      stat = "identity",
      position = "dodge"
    ) + 
    scale_fill_brewer(palette="Spectral") + 
    labs(
      x = "Brand",
      y = "Number of Ads"
    )
  
  # Return the plot objects
  print(plot)
}
```

Here are the most interesting items I found from these plots

1. Budweiser has a large number of ads using animals which is not
common across the board. This makes sense as they have their famous
clydesdales’ commercials which have become a staple in the Super
Bowl lineup
2. The food/beverage companies show their products quickly and try
to do so comedically. The one notable point here is that Pepsi
loves to use celebrities in its commercials. 
3. NFL stays away from commercials involving danger, sex, or
animals - for a professional organization, I’d say this checks out.
4. Surprisingly, a car brand, KIA, uses sex quite frequently in
their ads. 

As the final part of this subquestion, I want to look at how ads have changed
over time. To do this, we'll need to first group the ads by year and then
reproduce our `melt` method from before to generate a plot:
```{r}
# Group the columns by each category
group_category_by_year <- super_bowl_ads %>%
  group_by(year) %>%
  summarize(
    funny = sum(funny),
    show_product_quickly = sum(show_product_quickly),
    patriotic = sum(patriotic), 
    celebrity = sum(celebrity),
    danger = sum(danger),
    animals = sum(animals),
    use_sex = sum(use_sex))

# Convert tible to dataframe
group_category_by_year <- as.data.frame(group_category_by_year)

# Melt all the values from the dataframe to create a new one
melt_df_year <- melt(group_category_by_year, id.vars = 1)

# Generate a grouped barplot
ggplot(
  data = melt_df_year,
  aes(x = year, y = value))+ 
  geom_bar(
    aes(fill = variable),
    stat = "identity",
    position = "stack") + 
  scale_fill_brewer(palette="Spectral") +
  labs(
    title = "No Discernable Pattern Across Time",
    x = "Year",
    y = "Number of Ads"
  )
```

As noted in the title of the graph, there doesn't seem to be any discernable 
patterns for our top 10 brands. What I would say is more surprising is that 
we don't see the number of ads they produce increase over time.

---

### (2) | Ad Superlatives

The first ad superlative we can give is "most popular." To determine which 
has been the most popular ad, we will simply look at which ad has the greatest
number of views:
```{r}
super_bowl_ads %>%
  arrange(desc(view_count)) %>%
  slice(1:10) %>%
  ggplot() + 
  aes(x = view_count, y = reorder(title, +view_count)) + 
  scale_x_continuous(labels = label_number(suffix = " M", scale = 1e-6)) + 
  geom_bar(
    stat = "identity",
    fill = '#3388bd') + 
  labs(
    x = "Number of Views",
    y = "Video Title"
  )
```

Seems like we have a clear winner here: "Doritos Sling Baby" with a whopping
175 million views! The next highest ad is the Budweiser "Born the Hard Way" ad
with only a fraction of the views. Such a disparity makes me wonder if this is 
the best way to determine "popularity" but we really aren't offered any better
metric to determine this. So we'll take it for what it is and continue with some
other superlatives.

#### Most Liked and Disliked Super Bowl Commercials
To determine this, I want to create four ratios:

1. Likes to Views Ratio
2. Dislikes to Views Ratios
3. Likes to Total Reactions (Likes + Dislikes)
4. Weighted Likes 
5. Weighted Dislikes

There are observations which have either no likes, no dislikes or no views. 
Since we cannot calculate the ratios for those commercials, we will exclude them
from this portion of the analysis.

Let's do it!
```{r}
# Exclude any values that are zero
super_bowl_like_dislike <- super_bowl_ads %>%
  filter(like_count > 0, dislike_count > 0, view_count > 0)

# Create the new features 
super_bowl_like_dislike <- super_bowl_like_dislike %>%
  mutate(
    ratio_likes_views = like_count / view_count,
    ratio_dislike_views = dislike_count / view_count,
    total_reactions = like_count + dislike_count,
    ratio_likes_react = like_count / total_reactions,
    ratio_dislikes_react = 1 - ratio_likes_react,
    weight = view_count / sum(view_count),
    likes_weighted = ratio_likes_react * weight,
    dislikes_weighted = ratio_dislikes_react * weight
  )

# We can display the top 10 ads
super_bowl_like_dislike %>%
  arrange(desc(ratio_likes_react)) %>%
  head(10) %>%
  select(title, ratio_likes_react)
```

With this dataframe now setup, let's go ahead and make some plots to see what 
our results are:
```{r}
# Display the top 10 ratio ads based on raw ratio_likes_react:
top_10_ratio_likes <- super_bowl_like_dislike %>%
  arrange(desc(ratio_likes_react)) %>%
  head(10) %>%
  ggplot() + 
  aes(y=reorder(title, +ratio_likes_react), x=ratio_likes_react) + 
  geom_bar(stat="identity", fill = '#3388bd') + 
  labs(
    title = "Not Much Difference at the Top",
    x = "Likes to Reactions Ratio",
    y = "Ad Title"
  )

# display the top 10 ratio ads now using the weighted likes:
top_10_ratio_likes_weighted <- super_bowl_like_dislike %>%
  arrange(desc(likes_weighted)) %>%
  head(10) %>%
  ggplot() + 
  aes(y=reorder(title, +likes_weighted), x=likes_weighted) + 
  geom_bar(stat="identity", fill = '#3388bd') + 
  labs(
    title = "Doritos Sling Baby Unsurprisingly Top",
    x = "Weighted Likes by Like Count",
    y = "Ad Title"
  )

# Display the two graphs
top_10_ratio_likes / top_10_ratio_likes_weighted
```

Looking by the raw likes to reactions ratio, we don't have much of a difference.
This is driven mainly by ads that have very few likes. This is evident because
none of the top 10 ads in the top chart are found in the bottom chart which 
accounts for the number of likes for the ad. Speaking to the weighted ratio, 
it is of no surprise that **Doritos Sling Baby** is at the top due to the sheer
number of views, which naturally, gives leads to more likes.

To prove this final point, let's quickly plot `view_count` versus `like_count`:
```{r}
# This will make a full plot
plot_outliers <- ggplot(super_bowl_ads, aes(x = view_count, y = like_count)) +
  geom_point(color = '#3388bd') + 
  labs(
    title = "Unfiltered Results",
    x = "Number of Views",
    y = "Number of Likes"
  ) 

# Filter out outliers
plot_no_outliers <- super_bowl_ads %>%
  subset(view_count <= quantile(view_count, 0.95)) %>%
  ggplot( aes(x = view_count, y = like_count)) +
  geom_point(color = '#3388bd') + 
  labs(
    title = "95% Quantile Results",
    x = "Number of Views",
    y = "Number of Likes"
  ) + 
  theme(axis.title.y = element_blank())

(plot_outliers | plot_no_outliers)
```

Now, let's do a similar analysis but now with the dislikes:
```{r}
# Display the top 10 ratio ads based on raw ratio_likes_react:
top_10_ratio_dislikes <- super_bowl_like_dislike %>%
  arrange(desc(ratio_dislikes_react)) %>%
  head(10) %>%
  ggplot() + 
  aes(y=reorder(title, +ratio_dislikes_react), x=ratio_dislikes_react) + 
  geom_bar(stat="identity", fill = '#3388bd') + 
  labs(
    title = "Ad Meter was not a Crowd Favorite",
    x = "Dislikes to Reactions Ratio",
    y = "Ad Title"
  )

# display the top 10 ratio ads now using the weighted likes:
top_10_ratio_dislikes_weighted <- super_bowl_like_dislike %>%
  arrange(desc(dislikes_weighted)) %>%
  head(10) %>%
  ggplot() + 
  aes(y=reorder(title, +dislikes_weighted), x=dislikes_weighted) + 
  geom_bar(stat="identity", fill = '#3388bd') + 
  labs(
    title = "Doritos Wins Loses the Battle, Budweiser, the War!",
    x = "Weighted Likes by Like Count",
    y = "Ad Title"
  )

# Display the two graphs
top_10_ratio_dislikes / top_10_ratio_dislikes_weighted
```

Here, we actually have a better view of the most disliked ad which is 🥁 🥁 🥁

**Budweiser 2017 Super Bowl Commercial | "Born The Hard Way**

While it's raw dislikes ratio is not the highest, it is the only ad from the top
graph which also shows in the bottom graph (in fact it places second in the 
weighted dislikes graph). 

### (3) | Categories and Views

Here, we're curious about the relationship between the number of categories
an ad has and their popularity (which again, we can measure by the
number of views an ad has).

Let's first create a new column to represent the number of categories an ad has
and then look at the top 10 ads in terms of popularity.
```{r}
# Create a column representing no. of ads:
super_bowl_ads$count_categories <- rowSums(
  super_bowl_ads[c("funny", "show_product_quickly", "patriotic", "celebrity", 
                   "danger", "animals", "use_sex")]
  )    

# Now having created that column let's go ahead and visualize the top 10 
# ads and their consequent category count:
super_bowl_ads %>%
  arrange(desc(view_count)) %>%
  head(10) %>%
  ggplot() +
  aes(y = reorder(title, +view_count), x = count_categories) + 
  geom_bar(
    stat = "identity",
    fill = '#3388bd'
  ) + 
  labs(
    title = "Top 10 Ads Have at Least 2 categories",
    x = "Number of Categories",
    y = "Ad Title"
  )
  
# Another way to visualize this would be by using a boxplot, however because I'm
# going to only take 90% quantile to get a nicer view of the majority of the 
# data
super_bowl_ads %>%
  filter(view_count <= quantile(view_count, 0.9)) %>%
  ggplot(aes(x = factor(count_categories), y = view_count)) +
  geom_boxplot(color = '#3388bd') + 
  labs(
    title = "No Real Advantage to Number of Categories",
    x = "Number of Categories",
    y = "Number of Views"
  )


```

From these graphs, there's really no conclusive statement that can be made about
the number of categories an ad has. What we can say is that the top ads will 
have at least more than 1 category. 

Another way to look at this would be simply by the raw number of views each
ad category has. We can do so with the following plot:
```{r}
# Again, here I'm taking the 90% to deal with outliers
super_bowl_ads %>%
  filter(view_count <= quantile(view_count, 0.9)) %>%
  group_by(count_categories) %>%
  summarize(view_count = sum(view_count)) %>%
  ggplot(aes(x = count_categories, y = view_count)) + 
  geom_bar(stat = "identity", fill='3388bd') + 
  scale_x_continuous(breaks = round(seq(min(super_bowl_ads$count_categories), 
                                        max(super_bowl_ads$count_categories), 
                                        by = 1),1))
```

Follows a normal distribution, which I think is to be expected as most ads will
definitely incorporate several elements, but not too few or too many so as to
oversimplify or convolute the message.

## Conclusion

With that, I want to wrap up this notebook. Overall, I didn't find the data to 
be too interesting since at the end of the day the exciting columns are quite
subjective. Also, because this is already a subset of the "most popular" ads, 
we don't have a holistic view of super bowl ads and their performance. 

Nevertheless, these notebooks are simply to further develop my R skills and I 
can definitely say I feel like I've achieved that purpose here. Looking forward 
to the next one :) 


