---
title: "World Cup EDA"
output:
  html_document:
    df_print: paged
---

# Purpose
To deeper my understanding of R I want to try my hand at exploring different data sets in which I can apply my newly learned skills.

For this notebook, I'll look at the [FIFA world cup dataset](https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-11-29) ⚽️ from the Tidy Tuesdays

---
## Prep

First, let's load the libraries and data:
```{r libraries, message=FALSE}
library(tidyverse)
library(ggplot2)
library(tidytuesdayR) # library from which we'll retrieve the data
library(patchwork)

# Let's also set a theme for our plots
theme_set(theme_minimal())
```

There are two dataframes that we'll index to two separate dataframe objects
```{r data, message=FALSE}
world_cup <- tt_load('2022-11-29')
wc_matches <- world_cup$wcmatches
wc_cups <- world_cup$worldcups
```

## Data Review
Let's now take a quick look at both dataframes and see if they match expectations
```{r quick-view}
head(wc_matches)
head(wc_cups)
```

Furthermore, let's see if there is any empty data that we have to deal with:
```{r empty-data}
sum(is.na(wc_cups))
sum(is.na(wc_matches))
```

The `wc_matches` seems to have several missing values, let's review which
columns contribute the most to that number
```{r check-missing-values-columns}
sapply(wc_matches, function(x) sum(is.na(x)))
```

There are three columns that compose all the missing values, let's index those
and take a sample from each of those columns
```{r missing-values-contributors}
wc_matches %>%
  filter(!is.na(win_conditions)) %>%
  select(win_conditions) %>%
  head()

wc_matches %>%
  filter(is.na(winning_team)) %>%
  select(outcome, winning_team, losing_team) %>%
  head()
```

After reviewing this data we learned two important things:

  1. `NA` values in `win_conditions` occur when the game was either won, lost, or
drawn in regular time. Whenever this column is not empty, it means that the 
game was decided in extra time (`AET`) or in penalties
  2. The `winning_team` and `losing_team` both have the same amount of missing 
values. This should've been an indicator that `NA` values here represent draws.

Consequently, let's take the following action:

> Create a new column called `penalties` and a new column called `aet`. These
two columns will help to better characterize the results of the game

```{r feature-eng}
# This will only set the values to those columns that have a value for 
# win_conditions
wc_matches <- wc_matches %>%
  mutate(penalties = ifelse(str_detect(win_conditions, 'penalties'), 1, 0),
         aet = ifelse(str_detect(win_conditions, 'AET'), 1, 0))

# Set all the NA values in the two columns to zero
wc_matches[c("penalties", "aet")][is.na(wc_matches[c("penalties", "aet")])] <- 0
```

## EDA - World Cup Matches
Seeing as we've done most of the work on the `wc_matches` dataframe, we'll 
take a first swing at further investigating this data.

However, before we begin, let's ask some questions that will help to guide our
analysis:

  1. Who is the winningest/losingest/drawingest (not sure if that's a word) 
  team?
    - Review both raw wins and win %
  2. What team has played the most games in the WC? Is that associated with
  the having more wins? One would think yes
  3. What team has scored the most goals and has had the most goals scored 
  against them?
  4. How many matches have required extra time? If so, how many have required
  penalties?
  5. Do scores vary across dates? Meaning do we see more wins/losses/draws on a 
  certain day of the week?
  6. Does being home or away have any impact on the outcome of the game?
  7. Do we see a difference in the amount of goals scored year to year? Or even
  day to day?
  
For now, I think that list of questions should be enough to get us started : )

### 1. & 2. Winningest/Losingest/?Drawingest? Teams and Most Games

#### 1. Winningest/Losingest/?Drawingest? Teams
```{r win/lose/draw}

# Grouping the wins
wc_wins <- wc_matches %>%
  filter(!is.na(winning_team)) %>%
  group_by(winning_team) %>%
  summarize(wins=n()) %>%
  arrange(desc(wins)) %>%
  rename(team = winning_team)

# Grouping the losses
wc_losses <- wc_matches %>%
  filter(!is.na(losing_team)) %>%
  group_by(losing_team) %>%
  summarize(losses=n()) %>%
  arrange(desc(losses)) %>%
  rename(team = losing_team)

# Grouping the draws by home and away
wc_draws_home <- wc_matches %>%
  filter(is.na(win_conditions)) %>%
  group_by(home_team) %>%
  summarize(draws_home=n()) %>%
  arrange(desc(draws_home)) %>%
  rename(team = home_team)

wc_draws_away <- wc_matches %>%
  filter(is.na(win_conditions)) %>%
  group_by(away_team) %>%
  summarize(draws_away=n()) %>%
  arrange(desc(draws_away)) %>%
  rename(team = away_team)

# We then have to merge the two draws and add the results:
wc_draws <- inner_join(x = wc_draws_home, y = wc_draws_away)
wc_draws$draws <- wc_draws$draws_away + wc_draws$draws_home
wc_draws <- wc_draws[c("team", "draws")]

# We can then merge all the objects together
wc_results <- inner_join(wc_wins, wc_losses)
wc_results <- inner_join(wc_results, wc_draws)

# Then create a final column which adds up all the games played
wc_results$games <- rowSums(wc_results[c("wins", "losses", "draws")])
```

Now, we can go ahead and begin to visualize the different results from our games
```{r visualize-games}
par(mfrow=c(3,1))

graph_results <- function(df, x, y) {
  x = enquo(arg = x)
  y = enquo(arg = y)
  
  result_graph <- df %>%
    head(10) %>%
    ggplot() + 
    aes(x=reorder(!!x, -!!y), y=!!y) + 
    geom_bar(stat='identity') + 
    labs(
      x = x,
      y = y
    )
  
  return(result_graph)
}

wins <- graph_results(wc_results, team, wins)
losses <- graph_results(wc_results, team, losses)
draws <- graph_results(wc_results, team, draws)

wins / losses / draws
```

Brazil is the winningest and drawingest team?, Argentina is the losingest team.
However, if we take a look at winning percentage, do we still see a similar
arrangement?

```{r winning-pct}
wc_results$win_pct <- wc_results$wins / wc_results$games

wc_results %>% 
  head(10) %>%
  ggplot() + 
  aes(x=reorder(team, -win_pct), y=win_pct) + 
  geom_bar(stat='identity') + 
  labs(title = 'Brazil is still at the top!',
       x='Team',
       y='Win Percentage')
```

There is the topic of `Germany` and `West Germany` which I have ignored. Should
we count these two together or leave them separate? For now, I'll leave them be.


#### 2. Most Games Played
With the data already structure, it's quite easy to determine the team with the 
most games by calling the `games` column in the `wc_results`. 

So let's grab those results and also plot them agains the `wins` column to see
if there is any evident trend between games played and wins:
```{r games-played}
# Getting the top-5 countries with the most games played
wc_results %>%
  arrange(desc(games)) %>%
  head(5) %>%
  print()

# Plotting the games played vs wins
wc_results %>%
  ggplot() + 
  aes(x=wins, y=games) +
  geom_point(shape=1) + 
  labs(title = 'More games = More wins',
       x = 'Number of Wins',
       y = 'Number of Games Played')
```

From the scatterplot, we can clearly see that the more games a team has played, 
the more games they've won. This makes sense as if you get pass the group stages
of the world cup, you will get to play more games as you win.

### 3. What team has scored the most goals? 
```{r goals}
# Function to group the teams by home/away and home/away goals
wc_goals_fx <- function(df, group, col){
  group = enquo(arg = group)
  col = enquo(arg = col)
  
  wc_goal_type <- df %>%
    group_by(!!group) %>%
    summarize(sum(!!col)) %>%
    rename(team = !!group)
  
  return(wc_goal_type)
}

# Goals Scored For and Against, Home and Away
goals_home_sco <- wc_goals_fx(wc_matches, home_team, home_score)
goals_away_sco <- wc_goals_fx(wc_matches, away_team, away_score)
goals_home_con <- wc_goals_fx(wc_matches, home_team, away_score)
goals_away_con <- wc_goals_fx(wc_matches, away_team, home_score)

# Goals Scored
goals_scored <- inner_join(goals_home_sco, goals_away_sco)
goals_scored$total_goals_scored <- goals_scored$`sum(home_score)` + goals_scored$`sum(away_score)`
goals_scored <- goals_scored[c('team', 'total_goals_scored')]

# Goals Conceded
goals_conceded <- inner_join(goals_home_con, goals_away_con)
goals_conceded$total_goals_conceded <- goals_conceded$`sum(home_score)` + goals_conceded$`sum(away_score)`
goals_conceded <- goals_conceded[c('team', 'total_goals_conceded')]

# Join the two and clean up the dataframe
goals <- inner_join(x = goals_scored, y = goals_conceded)
goals$differential <- goals$total_goals_scored - goals$total_goals_conceded
goals <- rename(
  goals, 
  scored = total_goals_scored, 
  conceded = total_goals_conceded
)

# Let's do a quick scatterplot
scored_v_conceded <- goals %>%
  ggplot() + 
  aes(x = conceded, y = scored) +
  geom_point(shape = 1) + 
  labs(title = "More Goals Scored Means More Goals Conceded",
       x = "Goals Conceded",
       y = "Goals Scored")

# Additionally, let's look at the distribution of goal differentials:
goal_differential <- goals %>% 
  ggplot() + 
  aes(x=differential) + 
  geom_histogram(
    binwidth=5,
    colour="black",
    fill="white") + 
  geom_vline(
    xintercept=mean(goals$differential),
    linetype="dashed",
    color="red") + 
  labs(
    title = "Mostly Normal Distribution and Brazil...",
    x = "Goal Differential",
    y = "Count")

scored_v_conceded / goal_differential
```
